# Examples

Here we provide examples of using different aspects of the `conTree`
package as applied to several data sets. These examples are intended to
illustrate some of the basic features, operation and control of the
package procedures. See the full documentation for more advanced
options. In this note lines beginning with " $>$ " are to be interpreted
as R commands.

## Conditional probability estimation


Here we present R code and resulting output using `conTree` for the
analysis presented in Section 6.2 of @FriedmanPNAS using the census
income data. This sample, taken from 1994 US census data, consists of
observations from 48842 people divided into a training set of 32561 and
an independent test set of 16281. The binary outcome variable
$y\in\{0,1\}$ indicates whether or not a person's income is greater than
\$50000 per year. There are 14 predictor variables $\mathbf{x}$
consisting of various demographic and financial properties associated
with each person. The goal is to contrast $y$ with estimates of
$\Pr(y=1\,|\,\mathbf{x})$ obtained by several machine learning methods:
gradient boosting on logistic scale using maximum likelihood (GBL),
random forest (RF), and gradient boosting on the probability scale (GBP)
using least--squares.

First load `conTree` package

```{r}
library(`conTree`)
```

The next step is to load and attach the data with the R command

```{r}
data(census, package = "conTree")
attach(census)
```

This creates the following numeric vectors and data frames :

|Variable  |Description                                    |
|:---------|:----------------------------------------------|
|`x`       |training data predictor variables (data frame) |
|`y`       |training data outcome (high salary indicator)  |
|`xt`      |test data predictor variables (data frame)     |
|`yt`      |test data outcome                              |
|`gbl`     |training data GBL estimates                    |
|`gblt`    |test data GBL estimates                        |
|`rf`      |training data RF estimates                     |
|`rft`     |test data RF estimates                         |
|`gbp`     |training data GBP estimates                    |
|`gbpt`    |test data GBP estimates                        |



Note that all of the $\Pr(y=1\,|\,\mathbf{x})$ estimates were obtained
using the training data ($\mathtt{x}$,`y`) only.

Partition *test* data into two parts

```{r}
dx = 1:10000
dxt = 10001:16281
```

Contrast $y$ and GBL estimates using first part

```{r}
tree = contrast(xt[dx,], yt[dx], gblt[dx], type = 'prob')
```

Validate on second part which produces the command line output:

```{r}
nodesum(tree, xt[dxt,], yt[dxt], gblt[dxt])
```

The first component `$nodes` lists the tree node identifiers for each
terminal node in order of region discrepancy. The second `$cri` gives
the actual discrepancy of each corresponding region. The third component
`$wt` shows the corresponding number of observations in each region. The
quantity `$avecri` is the observation weighted average discrepancy over
all regions.

The function `nodeplots()`  produces the graphical tree summary shown
in Figure \@ref(fig:fig1).

```{r fig1, fig = TRUE, fig.cap = "Graphical tree summary."}
nodeplots(tree, xt[dxt,], yt[dxt], gblt[dxt])
```

The blue bars in the upper barplot represent the empirical $\Pr(y=1)$
in each region whereas the red ones show the mean of the GBL
predictions in the corresponding regions. One sees dramatic over
estimation of the small probabilities.

Region boundaries for selected nodes 7 and 29 can be obtained by:

```{r}
treesum(tree, c(7, 29))
```

Interpretation of this output is described in Section \@ref(treesum).

The command

```{r}
nx = getnodes(tree, xt)
```

obtains the terminal node identity for each test observation for the
tree represented in Figure \@ref(fig:fig1). The commands

```{r fig2, fig = TRUE, fig.cap = "GBL vrsus GBP probability estimates for test data observations in region 7."}
plot(gblt[nx == 7], gbpt[nx == 7], pch = '.', xlab = 'GBTL', ylab = 'GBPT')
lines(c(0, 1), c(0, 1), col = 'red')
```

plot the test data GBP predictions against those of GBL for the
highest discrepancy region 7. The red line represents equality. The
result shown in Figure \@ref(fig:fig2) indicates that the gradient
boosting probabilities based on log-odds estimates in this region are
considerably larger but proportional to those obtained by direct
probability estimation using least-squares. One also sees the effect
of truncating the out of range estimates produced by the latter. In
spite of this the GBP probability estimates are seen in Figure
\@ref(fig:fig3) (below) to be from three to four times more accurate
than those of GBL

We next plot the lack-of-fit contrast curve for GBL using a 50 terminal
node contrast tree

```{r, eval = FALSE}
tree = contrast(xt[dx,], yt[dx], gblt, type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt,], yt[dxt], gblt[dxt], doplot = 'first')
```

Next we add the corresponding curve for RF predicted probabilities

```{r, eval = FALSE}
tree = contrast(xt[dx, ], yt[dx], rft[dx, 2], type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt, ], yt[dxt], rft[dxt, 2], doplot = 'next', col = "blue")
```

And finally that for the GBP estimates

```{r, eval = FALSE}
tree = contrast(xt[dx, ], yt[dx], gbpt[dx], type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt, ], yt[dxt], gbpt[dxt], doplot = 'next', col = "red")
```

```{r fig3, echo = FALSE, fig = TRUE, fig.cap = "Lack-of-fit contrast curves for GBL (black), RF (blue), and GBP (red) for census income data."}
tree = contrast(xt[dx,], yt[dx], gblt, type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt,], yt[dxt], gblt[dxt], doplot = 'first')
tree = contrast(xt[dx, ], yt[dx], rft[dx, 2], type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt, ], yt[dxt], rft[dxt, 2], doplot = 'next', col = "blue")
tree = contrast(xt[dx, ], yt[dx], gbpt[dx], type = 'prob', tree.size = 50)
lofcurve(tree, xt[dxt, ], yt[dxt], gbpt[dxt], doplot = 'next', col = "red")
```

The result is shown in Figure \@ref(fig:fig3).


## Contrast boosting

Here we illustrate the use of contrast boosting to improve prediction
performance of GBL. The model is built using the *training* data with
the command

```{r}
modgbl = modtrast(x, y, gbl, type = 'prob', niter = 200)
```

We can plot tree average discrepancy versus iteration number on the
training data and super impose a corresponding plot based on the test
data set with the commands:

```{r fig4, fig = TRUE, fig.cap = "`xval(x, y, gbl, modgbl, col = 'red')`"}
xval(modgbl, x, y, gbl, col = 'red')
xval(modgbl, xt, yt, gblt, col = 'green', doplot = 'next')
```

The result is shown in Figure \@ref(fig:fig4). Note that results at
every tenth iteration are shown.

Model predictions on the test data set can be obtained with the command

```{r}
hblt = predtrast(modgbl, xt, gblt)
```

These can be contrasted with the $y$ - values using the first test set
sample

```{r}
tree = contrast(xt[dx, ], yt[dx], hblt[dx], type = 'prob')
```

and summarized on the other test sample with the command

```{r fig5, fig = TRUE, fig.cap = "`nodeplots(tree, xt[dxt, ], yt[dxt], hblt[dxt])`"}
nodeplots(tree, xt[dxt, ], yt[dxt], hblt[dxt])
```

producing Figure \@ref(fig:fig5).

Comparing with Figure \@ref(fig:fig5) one sees that contrast boosting
the GBL model appears to have substantially reduced the shrinking of
the of the extreme probabilities estimates thereby improving its
conditional probability estimation accuracy.

One can boost the RF and GBP models in the same way using the commands
analogous with those used here for the GBL model. Additionally,
lack-of-fit contrast curves can be produced and plotted for all three
boosted models using modifications of the corresponding commands above.
The results are shown in Fig. 4 of Friedman (2020).

## Conditional distributions

We illustrate contrasting and estimating conditional distributions
$p_{y}(y\,|\,\mathbf{x})$ using the demographics data set described in
Table 14.1 of Hastie, Tibshirani and Friedman (2008). Here we attempt to
estimate a persons age as a function of the other 13 demographic
variables. For this data set age value is reported as being in one of
seven intervals
$$age\in\text{\{13-17, 18-24, 25-34, 35-44, 45-54, 55-64,}\geq\text{ 65\}.}$$
As input to the algorithm each persons age is randomly generated
uniformly within its corresponding reported interval. For the last
interval an exponential distribution was used with mean corresponding to
life expectancy after reaching age 65.

We first load the data and attach it.

```{r}
data(age_data, package = "conTree")
attach(age_data)
```

This loads the following numeric vectors and data frames:

|Variable |Description                                            |
|:--------|:------------------------------------------------------|
|`xage`   |outcome variable (age)                                 |
|`yage`   |predictor variables (other demographics - data frame)  |
|`gbage`  |gradient boosting model for median (yage $\vert$ xage) |


The command

```{r fig6, fig = TRUE, fig.cap = "`hist(age)`"}
hist(yage)
```

produces the marginal age distribution as seen in Figure \@ref(fig:fig6).

This data is divided into training and test data subsets

```{r}
dl = 1:5000 # training data`
dt = 5001:8856 # test data`
```

We next contrast the distribution of $y\,|\,\mathbf{x}$ with that of
the no information hypothesis $p_{y}(y\,|\,\mathbf{x})=\hat{p}_{y}(y)$
where $\hat {p}_{y}(y)$ is the empirical marginal $y$ - distribution
independent of $\mathbf{x}$ as shown in Figure \@ref(fig:fig6). First
create independent marginal distribution of $y$ as the contrasting
distribution

```{r}
set.seed(5)
zage = yage[sample.int(length(yage))]
```

Then contrast with $y$ - distribution

```{r}
treezage = contrast(xage[dt, ], yage[dt], zage[dt], tree.size = 9, min.node = 200)
```

with tree test set command line summary produced by

```{r}
nodesum(treezage, xage[dt, ], yage[dt], zage[dt])
```

and graphical summary produced by

```{r fig7, fig = TRUE, fig.cap = "`nodeplots(treezage, xage[dt, ], yage[dt], zage[dt])`"}
nodeplots(treezage, xage[dt, ], yage[dt], zage[dt])
```

shown in Figure \@ref{fig:fig7).

Each frame in Figure \@ref{fig:fig7) shows a QQ-plot of the
distribution of age $y$ versus that of its $\mathbf{x}$
independent marginal counterpart (Figure \@ref(fig:fig6)) in each of
eight regions returned by contrast. To the extent the two
distributions are similar the points would lie close to the diagonal
(red) line. Here they are see to be very different indicating that
$p_{y}(y\,|\,\mathbf{x})$ is highly dependent on $\mathbf{x}$.

We next construct $y_{B}\sim p_{B}(y\,|\,\mathbf{x)}$, the residual
bootstrap approximation to $p_{y}(y\,|\,\mathbf{x)}$ using gradient
boosting median estimates for its location. This assumes that
$p_{y}(y\,|\,\mathbf{x)}$ is homoskedastic with varying location

```{r}
res = yage - gbage
```

```{r}
rbage = gbage + res[sample.int(length(res))]
```

and contrast $y$ with $y_{B}$ on the test data

```{r}
treerbage = contrast(xage[dt, ], yage[dt], rbage[dt], tree.size = 9, min.node = 200)
```

```{r fig8, fig = TRUE, fig.cap ="`nodeplots(treerbage, xage[dt, ], yage[dt], rbage[dt])`" }
nodeplots(treerbage, xage[dt, ], yage[dt], rbage[dt])
```

The result is shown in Figure \@ref{fig:fig8).

Although not perfect, the residual bootstrap $p_{B}(y\,|\,\mathbf{x)}$
is seen to provide a much closer approximation to
$p_{y}(y\,|\,\mathbf{x)}$ than the global marginal $p_{y}(y)$. This
indicates that its location has a strong dependence on $\mathbf{x}$ as
captured by the gradient boosting conditional median estimate.

Next we attempt to further improve the estimate of
$p_{y}(y\,|\,\mathbf{x)}$ by applying distribution boosting to the
training data starting with the residual boostrap approximation.

```{r}
mdlrb = modtrast(xage[dl, ], yage[dl], rbage[dl], min.node = 200)
```

The commands

```{r fig9, fig = TRUE, fig.cap = "`xval(mdlrb, xage[dl, ], yage[dl], rbage[dl], col = 'red')`"}
xval(mdlrb, xage[dl, ], yage[dl], rbage[dl], col = 'red')
xval(mdlrb, xage[dt, ], yage[dt], rbage[dt], col = 'green', doplot = 'next')
```

produce a plot of training (red) and test (green) data average tree
discrepancy as a function of iteration number (every 10th iteration) as
shown is Figure \@ref(fig:fig9).

The command

```{r}
hrbage = predtrast(mdlrb, xage[dt, ], rbage[dt])
```

transforms the test data residual bootstrap distribution
$y_{B}\sim p_{B}(y\,|\,\mathbf{x)}$ to the $y$ distribution estimate
$\hat{y}\sim\hat{p}_{\hat{y}}(y\,|\,\mathbf{x)}$.

The commands

```{r fig10, fig = TRUE, fig.cap = "`nodeplots(treerbage, xage[dt, ], yage[dt], hrbage)`"}
treehrbage = contrast(xage[dt, ], yage[dt], hrbage, tree.size = 9, min.node = 200)
nodeplots(treerbage, xage[dt, ], yage[dt], hrbage)
```

contrast the transformed distribution $\hat{y}$ with that of $y$ on the
test data. Results are shown in Figure \@ref(fig:fig10).

As seen the results are not perfect but somewhat better than that for
the residual bootstrap distribution shown in Figure \@ref(fig:fig8).
This is verified by the corresponding lack-of-fit contrast curves
shown in Figure \@ref(fig:fig11) as produced by the commands

```{r fig11, fig = TRUE, fig.cap = "Lack-of-fit contrast curves"}
lofcurve(treezage, xage[dt, ], yage[dt], zage[dt], doplot = 'first')
lofcurve(treerbage, xage[dt, ], yage[dt], rbage[dt], doplot = 'next', col = "blue")
lofcurve(treehrbage, xage[dt, ], yage[dt], hrbage, doplot = 'next', col = "red")
```

The lack-of-fit contrast curves are plotted for the three estimates of
$p_{y}(y\,|\,\mathbf{x})$: global marginal $p_{y}(y)$ (black), residual
bootstrap $p_{B}(y\,|\,\mathbf{x)}$ (blue) and distribution boosting
estimate $\hat{p}_{y}(y\,|\,\mathbf{x})$ (red). Distribution boosting is
seen to improve the accuracy of the conditional distribution estimate by
roughly a factor of two.

Finally we estimate $p_{y}(y\,|\,\mathbf{x})$ at $\mathbf{x}$ - values
for nine selected observations

```{r}
obs = c(8843, 5716, 7831, 6505, 4949, 7555, 3202, 6048, 7134)
```

at 500 quantile values

```{r}
p = ((1:500) -0.5) / 500
qres = as.numeric(quantile(res, p))
```

and plot their CDFs with

```{r fig12, fig = TRUE, fig.cap = "CDF estimates for nine $\\mathbf{x}$-values."}
opar <- par(mfrow=c(3,3))
for (k in 1:9) {
    plot(ydist(mdlrb, xage[obs[k], ], gbage[obs[k]] + qres), p, type = 'l', xlim=c(13, 100),
         xlab = 'Age', main = paste('Observation', as.character(obs[k])))
    points(yage[obs[k]], 0, col = 'red')
    title(paste('Observation', obs[k]))
}
par(opar)
```
as shown in Figure \@ref{fig:fig12). The red points shown at the bottom of each plot
display the actual realized $y$ - value (age) for that observation.
Prediction intervals for each observation can be read directly from its
corresponding CDF display.

Probability densities for these observations can be visualized with the
commands

```{r fig13, fig = TRUE, fig.cap = "Corresponding probability densities for the nine observations."}
opar <- par(mfrow = c(3,3))
for (k in 1:9) {
    hist(ydist(mdlrb, xage[obs[k], ], gbage[obs[k]] + qres), xlim = c(13, 100), nclass = 10,
         xlab = 'Age', main = paste('Observation', as.character(obs[k])))
    points(yage[obs[k]], 0, col = 'red')
    title(paste('Observation', obs[k]))
}
par(opar)
```
as seen in Figure \@ref(fig:fig13).

Considerable heteroskedasticity and skewness in the estimated
conditional distributions are evident.

## Two sample contrast trees

The use of two sample contrast trees is illustrated on the air quality
data set also from the Irvine Machine Learning Data Repository. The data
set consists of hourly averaged measurements from an array of 5 metal
oxide chemical sensors embedded in an air quality chemical multisensor
device. The outcome variable $y$ is  the corresponding true hourly
averaged concentration CO taken from a reference analyzer. The input
variables $\mathbf{x}$ are taken to be the corresponding hourly averaged
measurements of the other 13 quantities as described at the download web
site. The goal here is to contrast the conditional distribution of
$y|\,\mathbf{x}$ for data taken in the first six months (January --
June) to that of the last six months (July -- December).

The first step is to load the data

```{r}
data(air_quality)
attach(air_quality)
```

This loads in three vectors and a data frame

|Variable |Description                                            |
|:--------|:------------------------------------------------------|
|`yco`    |outcome variable (CO concentration)                    |
|`xco`    |predictor variables (data frame)                       |
|`zco`    |sample membership indicator                            |
|`pr2`    |probability propensity score.                          |

The first quantity `yco` is the outcome variable $y$, `xco` is a data
frame containing the 13 predictor variables $\mathbf{x}$ for each
observation and `zco` indicates sample membership: `zco < 0`
$\Rightarrow$ first six months, `zco > 0` $\Rightarrow$ last six months.
The vector `pr2` contains gradient boosting model predictions for
$\Pr(z>0\,|\,\mathbf{x})$; that is the predicted probability of each
observation belonging to the second sample as estimated from the
predictor variables.

We first contrast the means $E(y\,|\,\mathbf{x})$ of the different
samples as a function of $\mathbf{x}$. The command

```{r}
c(mean(yco[zco < 0]), mean(yco[zco > 0]))
```

displays the global means of the two samples with mean difference
$0.058$.

We first create learning and test data sets

```{r}
dl = 1:6000  # learning`
dt = 6001:9357 # test`
```

and contrast the means as a function of $\mathbf{x}$ as follows.

```{r}
tree = contrast(xco[dl, ], yco[dl], zco[dl], mode = 'twosamp', type = 'diffmean')
nodesum(tree, xco[dt,], yco[dt], zco[dt])
```
Here cri represents the mean difference in each of the nine regions
uncovered by the contrast tree. One sees that there are local regions of
the $\mathbf{x}$ - space where the mean difference between the two
samples is much larger than that of the global means.

Since the contrast tree regions are of finite extent mean differences
within each can originate from two sources. One is due to actual
differences in the conditional distributions $p_{y}(y\,|$
$\,\mathbf{x,\,}z<0)\,$ and $p_{y}(y\,|\,\mathbf{x,\,}z>0)$ in the
region. The other is differences in the marginal $\mathbf{x}$ -
distributions $p_{\mathbf{x}}(\mathbf{x}\,|\,z<0)$ and
$p_{\mathbf{x}}(\mathbf{x}\,|\,z>0)$ over each individual region. Since
interest is usually in the former one can mitigate the influence of the
latter by propensity weighting based on the propensity probability
scores. The command

```{r fig14, fig = TRUE, fig.cap = "`hist(pr2, nclass = 100)`"}
hist(pr2, nclass = 100)
```

displays the distribution of the input propensity probability scores
in Figure \@ref(fig:fig14). One sees that there are moderate
differences between the $\mathbf{x}$ - distributions of the two
samples at least globally.

The propensity weights are calculated with the commands

```{r}
wp = rep(0, length(zco))
wp[zco > 0] = 1 / pr2[zco > 0]
wp[zco < 0] = 1 / (1 - pr2[zco < 0])
wp = length(wp) * wp / sum(wp)
```

The corresponding (weighted) contrast tree is obtained as below.

```{r}
tree = contrast(xco[dl, ], yco[dl], zco[dl], w = wp[dl], mode = 'twosamp', type = 'diffmean')
nodesum(tree, xco[dt, ], yco[dt], zco[dt], w = wp[dt])
```

The component \$wt contains the sum of the weights in each region.
Although discrepancies remain, they are somewhat reduced when accounting
for the differences of the $\mathbf{x}$ - distributions within each
region. The command

```{r fig15, fig = TRUE, fig.cap = "`nodeplots(tree, xco[dt,], yco[dt], zco[dt], w = wp[dt])`"}
nodeplots(tree, xco[dt,], yco[dt], zco[dt], w = wp[dt])
```

produces the graphical summary shown in Figure \@ref(fig:fig15). The
 red/blue bars on the lower plot represent the fraction of the total
 weights of the first/second sample in each region.

A reference (null) distribution for two sample tree statistics under the
hypothesis of no difference between the samples can be obtained by
permutation testing. The analysis is repeatedly performed with randomly
permuted $z$ - values, thereby randomly assigning observations to each
sample. The commands

```{r fig16, fig = TRUE, fig.cap = "Two sample null distribution."}
avedisc = rep(0, 1000)
set.seed(13)
for (k in 1:1000) {
    zcot = zco[sample.int(length(zco))]
    tre = contrast(xco[dl, ], yco[dl], zcot[dl], mode = 'twosamp', type = 'diffmean')
    avedisc[k] = nodesum(tre, xco[dt,], yco[dt], zcot[dt])$avecri
}
hist(avedisc, xlim = c(0, 4))
points(c(2.937557, 3.790423), c(0, 0), col = c('blue', 'red'))
```
compute and display average tree discrepancy for 1000 replications of
this procedure. Figure \@ref(fig:fig16) shows the histogram of these null discrepancies along
with the corresponding propensity weighted/unweighted alternatives shown
as the blue/red points. One sees that the results based on the original
sample assignments are highly significant.

The command

```{r}
treesum(tree,c(11,25))
```

produces the command line output.

As described in Section \@ref(treesum) the output for each terminal node shows the
sequence of splits that produced its corresponding region. The names
corresponding to each predictor variable can be obtained with the
command

```{r}
names(xco)
```

so that the boundaries of nodes 11 and 25 are defined by

|Node  |Rule                                   |
|:-----|:--------------------------------------|
|`11`  |`RH>106 & AH>5232 & NMHC.GT >1112`     |
|`25`  |`106<=RH< 596 & AH<=4600 & NOx.GT<334` |


We next contrast the conditional distributions $y\,|\,\mathbf{x}$ of
the two samples. Figure \@ref(fig:fig17) shows a QQ-plot between the
respective $y$ global distributions on the test data obtained by the
commands

```{r fig17, fig = TRUE, fig.cap = "`qqplot(ycodt[zco[dt] < 0], ycodt[zco[dt] > 0], pch = '.')`"}
ycodt = yco[dt]
qqplot(ycodt[zco[dt] < 0], ycodt[zco[dt] > 0], pch = '.')
lines(c(0, 200), c(0, 200), col = 'red')
```

The global $y$ - distributions of the two samples are seen to be nearly
identical. The corresponding (propensity weighted) contrast tree results
are obtained with the commands

```{r}
tree = contrast(xco[dl, ], yco[dl], zco[dl], w = wp[dl], mode = 'twosamp', tree.size = 9)
nodeplots(tree, xco[dt,], yco[dt], zco[dt], w = wp[dt])
```

as shown in Figure \@ref{fig:fig16). The tree has uncovered a few
regions where there are moderate differences between the two
distributions.

## Session Information

```{r}
sessionInfo()
```
